Counterfactual Training: Teaching Models Plausible and Actionable Explanations

### 2025-07-23 paper/preprint/paper.tex
done

\begin{tcolorbox}[enhanced jigsaw, opacityback=0, bottomtitle=1mm, arc=.35mm, colback=white, coltitle=black, bottomrule=.15mm, breakable, leftrule=.75mm, title={Note \ref*{nte-train-default}: Training Phase}, toprule=.15mm, colframe=quarto-callout-note-color-frame, toptitle=1mm, opacitybacktitle=0.6, left=2mm, titlerule=0mm, rightrule=.15mm, colbacktitle=quarto-callout-note-color!10!white]

@@

\begin{tcolorbox}[enhanced jigsaw, opacityback=0, bottomtitle=1mm, arc=.35mm, colback=white, coltitle=black, bottomrule=.15mm, breakable, leftrule=.75mm, title={Note \ref*{nte-eval-default}: Evaluation Phase}, toprule=.15mm, colframe=quarto-callout-note-color-frame, toptitle=1mm, opacitybacktitle=0.6, left=2mm, titlerule=0mm, rightrule=.15mm, colbacktitle=quarto-callout-note-color!10!white]

@@

\subsubsection{Costs}\label{costs}

\caption{\label{tbl-costs}Costs}

@@

\begin{tabular}{lc}
  \toprule
  \textbf{Data} & \textbf{Cost $(-\%)$} \\\midrule
  LS & -26.82\pm0.86 $^{*}$ \\
  Circ & 40.97\pm0.82 $^{*}$ \\
  Moon & 33.83\pm0.98 $^{*}$ \\
  OL & 10.35\pm1.28 $^{*}$ \\\midrule
  Adult & 1.16\pm3.53 $^{}$ \\
  CH & -34.89\pm2.31 $^{*}$ \\
  Cred & 28.24\pm1.08 $^{*}$ \\
  GMSC & 3.54\pm5.78 $^{}$ \\
  MNIST & -31.67\pm7.72 $^{*}$ \\\midrule
  Avg. & 2.75 \\\bottomrule
\end{tabular}

@@

\subsubsection{Validity}\label{validity}

@@

\caption{\label{tbl-val}Validity}

@@

\caption{\label{tbl-val-mtbl}Validity}

@@

\begin{tcolorbox}[enhanced jigsaw, opacityback=0, bottomtitle=1mm, arc=.35mm, colback=white, coltitle=black, bottomrule=.15mm, breakable, leftrule=.75mm, title={Note \ref*{nte-gen-params-final-run-train}: Training Phase}, toprule=.15mm, colframe=quarto-callout-note-color-frame, toptitle=1mm, opacitybacktitle=0.6, left=2mm, titlerule=0mm, rightrule=.15mm, colbacktitle=quarto-callout-note-color!10!white]

@@

\begin{tcolorbox}[enhanced jigsaw, opacityback=0, bottomtitle=1mm, arc=.35mm, colback=white, coltitle=black, bottomrule=.15mm, breakable, leftrule=.75mm, title={Note \ref*{nte-gen-params-final-run-eval}: Evaluation Phase}, toprule=.15mm, colframe=quarto-callout-note-color-frame, toptitle=1mm, opacitybacktitle=0.6, left=2mm, titlerule=0mm, rightrule=.15mm, colbacktitle=quarto-callout-note-color!10!white]

@@

\begin{tcolorbox}[enhanced jigsaw, opacityback=0, bottomtitle=1mm, arc=.35mm, colback=white, coltitle=black, bottomrule=.15mm, breakable, leftrule=.75mm, title={Note \ref*{nte-pen-final-run-train}: Training Phase}, toprule=.15mm, colframe=quarto-callout-note-color-frame, toptitle=1mm, opacitybacktitle=0.6, left=2mm, titlerule=0mm, rightrule=.15mm, colbacktitle=quarto-callout-note-color!10!white]

@@

\begin{tcolorbox}[enhanced jigsaw, opacityback=0, bottomtitle=1mm, arc=.35mm, colback=white, coltitle=black, bottomrule=.15mm, breakable, leftrule=.75mm, title={Note \ref*{nte-pen-final-run-eval}: Evaluation Phase}, toprule=.15mm, colframe=quarto-callout-note-color-frame, toptitle=1mm, opacitybacktitle=0.6, left=2mm, titlerule=0mm, rightrule=.15mm, colbacktitle=quarto-callout-note-color!10!white]

@@

\begin{tcolorbox}[enhanced jigsaw, opacityback=0, bottomtitle=1mm, arc=.35mm, colback=white, coltitle=black, bottomrule=.15mm, breakable, leftrule=.75mm, title={Note \ref*{nte-train-final-run-train}: Training Phase}, toprule=.15mm, colframe=quarto-callout-note-color-frame, toptitle=1mm, opacitybacktitle=0.6, left=2mm, titlerule=0mm, rightrule=.15mm, colbacktitle=quarto-callout-note-color!10!white]

@@

\begin{tcolorbox}[enhanced jigsaw, opacityback=0, bottomtitle=1mm, arc=.35mm, colback=white, coltitle=black, bottomrule=.15mm, breakable, leftrule=.75mm, title={Note \ref*{nte-train-final-run-eval}: Evaluation Phase}, toprule=.15mm, colframe=quarto-callout-note-color-frame, toptitle=1mm, opacitybacktitle=0.6, left=2mm, titlerule=0mm, rightrule=.15mm, colbacktitle=quarto-callout-note-color!10!white]

@@

\begin{tcolorbox}[enhanced jigsaw, opacityback=0, bottomtitle=1mm, arc=.35mm, colback=white, coltitle=black, bottomrule=.15mm, breakable, leftrule=.75mm, title={Note \ref*{nte-tune-train}: Training Phase}, toprule=.15mm, colframe=quarto-callout-note-color-frame, toptitle=1mm, opacitybacktitle=0.6, left=2mm, titlerule=0mm, rightrule=.15mm, colbacktitle=quarto-callout-note-color!10!white]

@@

\begin{tcolorbox}[enhanced jigsaw, opacityback=0, bottomtitle=1mm, arc=.35mm, colback=white, coltitle=black, bottomrule=.15mm, breakable, leftrule=.75mm, title={Note \ref*{nte-tune-eval}: Evaluation Phase}, toprule=.15mm, colframe=quarto-callout-note-color-frame, toptitle=1mm, opacitybacktitle=0.6, left=2mm, titlerule=0mm, rightrule=.15mm, colbacktitle=quarto-callout-note-color!10!white]

@@

\begin{tcolorbox}[enhanced jigsaw, opacityback=0, bottomtitle=1mm, arc=.35mm, colback=white, coltitle=black, bottomrule=.15mm, breakable, leftrule=.75mm, title={Note \ref*{nte-tune_lr-train}: Training Phase}, toprule=.15mm, colframe=quarto-callout-note-color-frame, toptitle=1mm, opacitybacktitle=0.6, left=2mm, titlerule=0mm, rightrule=.15mm, colbacktitle=quarto-callout-note-color!10!white]

@@

\begin{tcolorbox}[enhanced jigsaw, opacityback=0, bottomtitle=1mm, arc=.35mm, colback=white, coltitle=black, bottomrule=.15mm, breakable, leftrule=.75mm, title={Note \ref*{nte-tune_lr-eval}: Evaluation Phase}, toprule=.15mm, colframe=quarto-callout-note-color-frame, toptitle=1mm, opacitybacktitle=0.6, left=2mm, titlerule=0mm, rightrule=.15mm, colbacktitle=quarto-callout-note-color!10!white]

>>>>

\begin{tcolorbox}[enhanced jigsaw, toptitle=1mm, toprule=.15mm, colback=white, colbacktitle=quarto-callout-note-color!10!white, leftrule=.75mm, arc=.35mm, opacitybacktitle=0.6, colframe=quarto-callout-note-color-frame, title={Note \ref*{nte-train-default}: Training Phase}, coltitle=black, titlerule=0mm, rightrule=.15mm, bottomtitle=1mm, breakable, bottomrule=.15mm, left=2mm, opacityback=0]

@@

\begin{tcolorbox}[enhanced jigsaw, toptitle=1mm, toprule=.15mm, colback=white, colbacktitle=quarto-callout-note-color!10!white, leftrule=.75mm, arc=.35mm, opacitybacktitle=0.6, colframe=quarto-callout-note-color-frame, title={Note \ref*{nte-eval-default}: Evaluation Phase}, coltitle=black, titlerule=0mm, rightrule=.15mm, bottomtitle=1mm, breakable, bottomrule=.15mm, left=2mm, opacityback=0]

@@

\begin{minipage}{0.50\linewidth}

\subsubsection{Integrated Gradients}\label{integrated-gradients}

\begin{table}[H]

\caption{\label{tbl-ig}Integrated gradients.}

@@

\begin{tabular}{
l
S[table-format=2.2(3.2)]
S[table-format=2.2(2.2)]
}
  \toprule
  \textbf{Data} & \textbf{CT} & \textbf{BL} \\\midrule
  LS & 0.03 \pm 0.0 & 10.24 \pm 2.4 \\
  Circ & 3.2 \pm 0.67 & 149.76 \pm 842.75 \\
  Moon & 60.84 \pm 128.51 & 0.55 \pm 0.06 \\
  OL & 0.78 \pm 0.12 & 4.81 \pm 1.08 \\\midrule
  Adult & 0.43 \pm 0.01 & 1.0 \pm 0.0 \\
  CH & 0.08 \pm 0.01 & 0.23 \pm 0.01 \\
  Cred & 0.0 \pm 0.0 & 0.43 \pm 0.01 \\
  GMSC & 1.0 \pm 0.0 & 0.21 \pm 0.03 \\
  MNIST & 0.18 \pm 0.01 & 0.41 \pm 0.01 \\
\end{tabular}

@@

\end{minipage}%
%
\begin{minipage}{0.50\linewidth}

\pandocbounded{\includegraphics[keepaspectratio]{../../paper/figures/mnist_ig.png}}

\caption{\label{fig-mnist-ig}Class-conditional integrated gradients.}

\subsubsection{Costs and Validity}\label{costs-and-validity}

@@

\caption{\label{tbl-panel}Costs and validity.}

\begin{minipage}{0.33\linewidth}

\subcaption{\label{tbl-costs}Costs}

\begin{tabular}{
l
S[table-format=2.2(1.2)]
}
  \toprule
  \textbf{Data} & \textbf{Cost $(-\%)$} \\\midrule
  LS & -26.82\pm0.86 $^{*}$ \\
  Circ & 40.97\pm0.82 $^{*}$ \\
  Moon & 33.83\pm0.98 $^{*}$ \\
  OL & 10.35\pm1.28 $^{*}$ \\\midrule
  Adult & 1.16\pm3.53 $^{}$ \\
  CH & -34.89\pm2.31 $^{*}$ \\
  Cred & 28.24\pm1.08 $^{*}$ \\
  GMSC & 3.54\pm5.78 $^{}$ \\
  MNIST & -31.67\pm7.72 $^{*}$ \\\midrule
  Avg. & 2.75 \\\bottomrule
\end{tabular}

\end{minipage}%
%
\begin{minipage}{0.33\linewidth}

\subcaption{\label{tbl-val}Validity}

@@

\end{minipage}%
%
\begin{minipage}{0.33\linewidth}

\subcaption{\label{tbl-val-mtbl}Validity}

@@

\begin{tcolorbox}[enhanced jigsaw, toptitle=1mm, toprule=.15mm, colback=white, colbacktitle=quarto-callout-note-color!10!white, leftrule=.75mm, arc=.35mm, opacitybacktitle=0.6, colframe=quarto-callout-note-color-frame, title={Note \ref*{nte-gen-params-final-run-train}: Training Phase}, coltitle=black, titlerule=0mm, rightrule=.15mm, bottomtitle=1mm, breakable, bottomrule=.15mm, left=2mm, opacityback=0]

@@

\begin{tcolorbox}[enhanced jigsaw, toptitle=1mm, toprule=.15mm, colback=white, colbacktitle=quarto-callout-note-color!10!white, leftrule=.75mm, arc=.35mm, opacitybacktitle=0.6, colframe=quarto-callout-note-color-frame, title={Note \ref*{nte-gen-params-final-run-eval}: Evaluation Phase}, coltitle=black, titlerule=0mm, rightrule=.15mm, bottomtitle=1mm, breakable, bottomrule=.15mm, left=2mm, opacityback=0]

@@

\begin{tcolorbox}[enhanced jigsaw, toptitle=1mm, toprule=.15mm, colback=white, colbacktitle=quarto-callout-note-color!10!white, leftrule=.75mm, arc=.35mm, opacitybacktitle=0.6, colframe=quarto-callout-note-color-frame, title={Note \ref*{nte-pen-final-run-train}: Training Phase}, coltitle=black, titlerule=0mm, rightrule=.15mm, bottomtitle=1mm, breakable, bottomrule=.15mm, left=2mm, opacityback=0]

@@

\begin{tcolorbox}[enhanced jigsaw, toptitle=1mm, toprule=.15mm, colback=white, colbacktitle=quarto-callout-note-color!10!white, leftrule=.75mm, arc=.35mm, opacitybacktitle=0.6, colframe=quarto-callout-note-color-frame, title={Note \ref*{nte-pen-final-run-eval}: Evaluation Phase}, coltitle=black, titlerule=0mm, rightrule=.15mm, bottomtitle=1mm, breakable, bottomrule=.15mm, left=2mm, opacityback=0]

@@

\begin{tcolorbox}[enhanced jigsaw, toptitle=1mm, toprule=.15mm, colback=white, colbacktitle=quarto-callout-note-color!10!white, leftrule=.75mm, arc=.35mm, opacitybacktitle=0.6, colframe=quarto-callout-note-color-frame, title={Note \ref*{nte-train-final-run-train}: Training Phase}, coltitle=black, titlerule=0mm, rightrule=.15mm, bottomtitle=1mm, breakable, bottomrule=.15mm, left=2mm, opacityback=0]

@@

\begin{tcolorbox}[enhanced jigsaw, toptitle=1mm, toprule=.15mm, colback=white, colbacktitle=quarto-callout-note-color!10!white, leftrule=.75mm, arc=.35mm, opacitybacktitle=0.6, colframe=quarto-callout-note-color-frame, title={Note \ref*{nte-train-final-run-eval}: Evaluation Phase}, coltitle=black, titlerule=0mm, rightrule=.15mm, bottomtitle=1mm, breakable, bottomrule=.15mm, left=2mm, opacityback=0]

@@

\begin{tcolorbox}[enhanced jigsaw, toptitle=1mm, toprule=.15mm, colback=white, colbacktitle=quarto-callout-note-color!10!white, leftrule=.75mm, arc=.35mm, opacitybacktitle=0.6, colframe=quarto-callout-note-color-frame, title={Note \ref*{nte-tune-train}: Training Phase}, coltitle=black, titlerule=0mm, rightrule=.15mm, bottomtitle=1mm, breakable, bottomrule=.15mm, left=2mm, opacityback=0]

@@

\begin{tcolorbox}[enhanced jigsaw, toptitle=1mm, toprule=.15mm, colback=white, colbacktitle=quarto-callout-note-color!10!white, leftrule=.75mm, arc=.35mm, opacitybacktitle=0.6, colframe=quarto-callout-note-color-frame, title={Note \ref*{nte-tune-eval}: Evaluation Phase}, coltitle=black, titlerule=0mm, rightrule=.15mm, bottomtitle=1mm, breakable, bottomrule=.15mm, left=2mm, opacityback=0]

@@

\begin{tcolorbox}[enhanced jigsaw, toptitle=1mm, toprule=.15mm, colback=white, colbacktitle=quarto-callout-note-color!10!white, leftrule=.75mm, arc=.35mm, opacitybacktitle=0.6, colframe=quarto-callout-note-color-frame, title={Note \ref*{nte-tune_lr-train}: Training Phase}, coltitle=black, titlerule=0mm, rightrule=.15mm, bottomtitle=1mm, breakable, bottomrule=.15mm, left=2mm, opacityback=0]

@@

\begin{tcolorbox}[enhanced jigsaw, toptitle=1mm, toprule=.15mm, colback=white, colbacktitle=quarto-callout-note-color!10!white, leftrule=.75mm, arc=.35mm, opacitybacktitle=0.6, colframe=quarto-callout-note-color-frame, title={Note \ref*{nte-tune_lr-eval}: Evaluation Phase}, coltitle=black, titlerule=0mm, rightrule=.15mm, bottomtitle=1mm, breakable, bottomrule=.15mm, left=2mm, opacityback=0]
